---
title: "Title"
output: html_document
echo: FALSE
---
## Executive Summary

### Loading the Data
The data is available from mtcars dataset. It is loaded using the data function. 
```{r load data, cache=TRUE}
data(mtcars)
```

### Exploratory Analysis

We create two boxplots to examine the following relationships:

* mpg v/s number of cylinders
* mpg v/s transimission type

Figure 1 in the appendix shows the plots. From the plot, it is clear that a relationship exists between mpg and number of cylinders and transmission type.  

### Regression Models
### Results

### Apendix
<figure>
```{r box plot, echo=FALSE}
par(mfrow=c(1,2))
with(mtcars, 
{
         boxplot(mpg~cyl,col="green4")
         title(main="Mpg vs number of cylinders", xlab = "Number of cylinders", ylab="miles per gallon")
         boxplot(mpg~am,col="blue", xaxt="n")
         title(main="Mpg vs transmission", xlab = "Transimission", ylab="miles per gallon")        
         axis(1, at=1:2, labels=c("Automatic", "Manual"))
})
```
<figcaption> Figure 1: Exploratory analysis of mpg v/s number of cylinders and transmission type <figcaption/>
</figure>